\centerline{Sudoku}

### 2021-05-25 report/paper.tex
oiafpbiuaspibfsa

    \setcounter{row}{1}
    \foreach \i in {1,...,9}
    {
    \setrow {$c_{\i,1}$ }{$c_{\i,2}$}{$c_{\i,3}$}  {$c_{\i,4}$}{$c_{\i,5}$}{$c_{\i,6}$}  {$c_{\i,7}$}{$c_{\i,8}$}{$c_{\i,9}$}
    }
  \end{scope}
\end{tikzpicture}
\end{center}
\caption{A $9\times 9$ Sudoku puzzle.}
\end{figure}
\noindent
Assuming that every $c_{i,j}$ in the Sudoku grid in Figure 1 takes just a value in interval \{1,2,3,4,5,6,7,8,9\}, such solution can be considered valid if and only if the following conditions are satisfied:
\begin{enumerate}[label={(\arabic*)}]
\item In every row $i$ (i.e., $c_{i,1}$, $c_{i,2}$, ..., $c_{i,9}$) there is at least one element that takes one value in \{1,2,3,4,5,6,7,8,9\}, but all values from 1 to 9 must be taken. Since in a row there are only 9 elements, it means that there will be a one-to-one mapping from these row elements to the \{1,2,3,4,5,6,7,8,9\} set values, thus each row will be made of nine different numbers.
\item In every column $j$ (i.e., $c_{1,j}$, $c_{2,j}$, ..., $c_{9,j}$) there is at least one element that takes one value in \{1,2,3,4,5,6,7,8,9\}, but all values from 1 to 9 must be taken. Since in a column there are only 9 elements, it means that there will be a one-to-one mapping from these column elements to the \{1,2,3,4,5,6,7,8,9\} set values, thus each column will be made of nine different numbers.
\item In every box, no number may appear more than once. Since a box is made up by 9 elements, it means that each box will contain all the values within the \{1,2,3,4,5,6,7,8,9\} set. What requires a bit more of reasoning here is finding a generalizing way to define all the elements in a box for every box. While for rows and columns it suffices to use a matrix-like notation, here some background reasoning is needed. It can be noticed that, in every box, the following pattern can be observed:
\begin{figure}[h]
\begin{center}
\begin{tikzpicture}[scale=.7]
  	\begin{scope}
    	\draw (0, 0) grid (3, 3);
   	\draw[very thick, scale=3] (0, 0) grid (1, 1);

@@

\noindent
It can be seen that the coordinates are nothing but all the possible permutations of three values: 1, 4 and 7. From this, we can define a set $B=\{1,\,4,\,7\}$ containing these values.\\
\end{enumerate}
\newpage
\section{Encoder and Decoder}\label{encdec}
\noindent
Let \textbf{is\_valid} define whether a row, column or box is valid. That means that the nine elements along a row, column or within a box must be different and equal to a digit in set $\{1,2,3,4,5,6,7,8,9\}$, thus meaning that a row, column or box is valid if it contains all digits. Let's consider $c_1,\,c_2,\,...,\,c_9$, which may be the elements within a row, column or box. Then we have:
\begin{align*}
\text{is\_valid}\left(  c_1,  c_2,  c_3,  c_4,  c_5,  c_6,  c_7,  c_8,  c_9
	\right) \Leftrightarrow & \\
& \left(  c_1  = 1 \vee  c_2  = 1 \vee c_3  = 1 \vee c_4  = 1 \vee  c_5  = 1
	\vee  c_6  = 1 \vee  c_7  = 1 \vee  c_8  = 1 \vee  c_9 = 1 \right)\\
& \wedge \left(  c_1  = 2 \vee  c_2  = 2 \vee c_3  = 2 \vee c_4  = 2 \vee  c_5
	= 2 \vee  c_6  = 2 \vee  c_7  = 2 \vee  c_8  = 2 \vee  c_9 = 2  \right)\\
& \wedge \left(  c_1  = 3 \vee  c_2  = 3 \vee c_3  = 3 \vee c_4  = 3 \vee  c_5
	= 3 \vee  c_6  = 3 \vee  c_7  = 3 \vee  c_8  = 3 \vee  c_9 = 3 \right)\\
& \wedge \left(  c_1  = 4 \vee  c_2  = 4 \vee c_3  = 4 \vee c_4  = 4 \vee  c_5
	= 4 \vee  c_6  = 4 \vee  c_7  = 4 \vee  c_8  = 4 \vee  c_9 = 4 \right)\\
& \wedge \left(  c_1  = 5 \vee  c_2  = 5 \vee c_3  = 5 \vee c_4  = 5 \vee  c_5
	= 5 \vee  c_6  = 5 \vee  c_7  = 5 \vee  c_8  = 5 \vee  c_9 = 5 \right)\\
& \wedge \left(  c_1  = 6 \vee  c_2  = 6 \vee c_3  = 6 \vee c_4  = 6 \vee  c_5
	= 6 \vee  c_6  = 6 \vee  c_7  = 6 \vee  c_8  = 6 \vee  c_9 = 6 \right)\\
& \wedge \left(  c_1  = 7 \vee  c_2  = 7 \vee c_3  = 7 \vee c_4  = 7 \vee  c_5
	= 7 \vee  c_6  = 7 \vee  c_7  = 7 \vee  c_8  = 7 \vee  c_9 = 7 \right)\\
& \wedge \left(  c_1  = 8 \vee  c_2  = 8 \vee c_3  = 8 \vee c_4  =  8 \vee  c_5
	= 8 \vee  c_6  = 8 \vee  c_7  = 8 \vee  c_8  = 8 \vee  c_9 = 8 \right)\\
& \wedge \left(  c_1  = 9 \vee  c_2  = 9 \vee c_3  = 9 \vee c_4  = 9 \vee  c_5
	= 9 \vee  c_6  = 9 \vee  c_7  = 9 \vee  c_8  = 9 \vee  c_9 = 9  \right)\\
\Leftrightarrow & \bigwedge^9_{n=1}  \bigvee^9_{i=1} \,\,\, c_i = n
\end{align*}
\hspace{0cm}\\
\noindent
This definition follows by the pigeon hole principle. Our aim is to check if the
nine inputs of \texttt{is\_valid} contain each one digit in
$\{1,2,3,4,5,6,7,8,9\}$. Each of the OR-clauses enclosed in brackets check if
each digit is present at least one time, but since we have 9 clauses it follows
that if even a single duplicate digit is given then at least an OR-clause will
have false value. If we imagine each OR-clause is a ``hole'' in which we deposit
digit, in this case the ``hole'' corresponding to the duplicate digit will be
``fed'' twice, and since we have only 9 ``holes'' and 9 ``pigeons'', the
duplicate will cause a ``hole'' to be void of ``pigeons'', thus making the
entire \texttt{is\_valid} formula false and correctly rejecting the invalid
input.
\hspace{0cm}\\

@@

\setcounter{equation}{0}
$\text{sudoku}\left( \{ c_{i,j} \}_{i, j \in \{ 1,2,3,4,5,6,7,8,9\}} \right) = $\begin{align}
&\bigwedge^9_{i=1} \text{ is\_valid}\left(  c_{i,1},  c_{i,2},  c_{i,3},  c_{i,4},  c_{i,5},  c_{i,6},  c_{i,7},  c_{i,8},  c_{i,9}  \right)\\
\wedge &\bigwedge^9_{j=1} \text{ is\_valid}\left( c_{1,j},  c_{2,j},  c_{3,j},  c_{4,j},  c_{5,j},  c_{6,j},  c_{7,j},  c_{8,j},  c_{9,j} \right) \\
	\wedge &\bigwedge^9_{i,j \in \{1,\,4,\,7\}} \text{ is\_valid}\left( c_{i,1},
	c_{i,(j+1)},  c_{i,(j+2)}, c_{(i+1),j},  c_{(i+1),(j+1)},
	c_{(i+1), (j+2)}, c_{(i+2),j},  c_{(i+2),(j+1)},  c_{(i+2),(j+2)} \right).
\end{align}
\newpage
\section{Development}\label{development}
We have developed the project in Python3 to provide both a web interface and the solver logic using z3.\\
For what concerns the z3 SAT solver, we have used the official \texttt{z3–solver} Python3 library to write the z3 logic instead of writing a z3 \texttt{.dimacs} file for every execution and invoking the z3 solver using a system shell. This allows for better performance and easier maintenance of the code.
The \texttt{z3–solver} Python3 library allows to specify constraints programmatically and is able to provide both information about the satisfiability information and data structure models that represent a valid solution for the problem.\\ \\
First of all, we want to translate the following definition into Python3 \texttt{z3–solver} code:
\begin{align*}
\text{sudoku}\left( \{ c_{i,j} \}_{i, j \in \{ 1,...,9\}} \right) = & \nonumber\\
       & \bigwedge^9_{i=1} \text{ is\_valid}\left(  c_{i,1},  c_{i,2},  c_{i,3},  c_{i,4},  c_{i,5},  c_{i,6},  c_{i,7},  c_{i,8},  c_{i,9}  \right)\\
\wedge &\bigwedge^9_{j=1} \text{ is\_valid}\left( c_{1,j},  c_{2,j},  c_{3,j},  c_{4,j},  c_{5,j},  c_{6,j},  c_{7,j},  c_{8,j},  c_{9,j} \right) \\
\wedge &\bigwedge^9_{i,j \in B} \text{ is\_valid}\left( c_{i,1},  c_{i,(j+1)},  c_{i,3},  c_{i,4},  c_{i,5},  c_{i,6},  c_{i,7},  c_{i,8},  c_{i,9} \right).
\end{align*}

@@

\begin{align*}
\text{is\_valid}\left(  c_1,  c_2,  c_3,  c_4,  c_5,  c_6,  c_7,  c_8,  c_9
	\right) = \bigwedge^9_{i=1}  \bigvee^9_{i=1} \,\,\, c_i
\end{align*}

>>>>

    \setcounter{row}{1}
    \foreach \i in {1,2,3,4,5,6,7,8,9}
    {
    \setrow {$c_{\i,1}$ }{$c_{\i,2}$}{$c_{\i,3}$}  {$c_{\i,4}$}{$c_{\i,5}$}{$c_{\i,6}$}  {$c_{\i,7}$}{$c_{\i,8}$}{$c_{\i,9}$}
    }
  \end{scope}
\end{tikzpicture}
\end{center}
\caption{A $9\times 9$ Sudoku puzzle.}
\end{figure}
\noindent
Assuming that every $c_{i,j}$ in the Sudoku grid in Figure 1 takes just a value in interval \{1,2,3,4,5,6,7,8,9\}, such solution can be considered valid if and only if the following conditions are satisfied:
\begin{enumerate}[label={(\arabic*)}]
\item In every row $i$ (i.e., $c_{i,1}$, $c_{i,2}$, $c_{i,3}$, $c_{i,4}$, $c_{i,5}$, $c_{i,6}$, $c_{i,7}$, $c_{i,8}$, $c_{i,9}$) there is at least one element that takes one value in \{1,2,3,4,5,6,7,8,9\}, but all values from 1 to 9 must be taken. Since in a row there are only 9 elements, it means that there will be a one-to-one mapping from these row elements to the \{1,2,3,4,5,6,7,8,9\} set values, thus each row will be made of nine different numbers.
\item In every column $j$ (i.e., $c_{1,j}$, $c_{2,j}$, $c_{3,j}$, $c_{4,j}$, $c_{5,j}$, $c_{6,j}$, $c_{7,j}$, $c_{8,j}$, $c_{9,j}$) there is at least one element that takes one value in \{1,2,3,4,5,6,7,8,9\}, but all values from 1 to 9 must be taken. Since in a column there are only 9 elements, it means that there will be a one-to-one mapping from these column elements to the \{1,2,3,4,5,6,7,8,9\} set values, thus each column will be made of nine different numbers.
\item In every box, no number may appear more than once. Since a box is made up by 9 elements, it means that each box will contain all the values within the \{1,2,3,4,5,6,7,8,9\} set. What requires a bit more of reasoning here is finding a generalizing way to define all the elements in a box for every box. While for rows and columns it suffices to use a matrix-like notation, here some background reasoning is needed. It can be noticed that, in every box, the following pattern can be observed:
\begin{figure}[h]
\begin{center}
\begin{tikzpicture}[scale=.7]
  	\begin{scope}
    	\draw (0, 0) grid (3, 3);
   	\draw[very thick, scale=3] (0, 0) grid (1, 1);

@@

\noindent
It can be seen that the coordinates are nothing but all the possible permutations of three values: 1, 4 and 7. From this, we can define a set $B=\{1,\,4,\,7\}$ containing these values.\\
\end{enumerate}
\newpage
\section{Encoder and Decoder}\label{encdec}
\noindent
Let \textbf{is\_valid} define whether a row, column or box is valid. That means that the nine elements along a row, column or within a box must be different and equal to a digit in set $\{1,2,3,4,5,6,7,8,9\}$, thus meaning that a row, column or box is valid if it contains all digits.
Let's consider $c_1,\,c_2,\,c_3,\,c_4,\,c_5,\,c_6,\,c_7,\,c_8,\,c_9$, which may be the elements within a row, column or box. Then we have:
\begin{align*}
\text{is\_valid}\left(  c_1,  c_2,  c_3,  c_4,  c_5,  c_6,  c_7,  c_8,  c_9
	\right) \Leftrightarrow & \\
& \left(  c_1  = 1 \vee  c_2  = 1 \vee c_3  = 1 \vee c_4  = 1 \vee  c_5  = 1
	\vee  c_6  = 1 \vee  c_7  = 1 \vee  c_8  = 1 \vee  c_9 = 1 \right)\\
& \wedge \left(  c_1  = 2 \vee  c_2  = 2 \vee c_3  = 2 \vee c_4  = 2 \vee  c_5
	= 2 \vee  c_6  = 2 \vee  c_7  = 2 \vee  c_8  = 2 \vee  c_9 = 2  \right)\\
& \wedge \left(  c_1  = 3 \vee  c_2  = 3 \vee c_3  = 3 \vee c_4  = 3 \vee  c_5
	= 3 \vee  c_6  = 3 \vee  c_7  = 3 \vee  c_8  = 3 \vee  c_9 = 3 \right)\\
& \wedge \left(  c_1  = 4 \vee  c_2  = 4 \vee c_3  = 4 \vee c_4  = 4 \vee  c_5
	= 4 \vee  c_6  = 4 \vee  c_7  = 4 \vee  c_8  = 4 \vee  c_9 = 4 \right)\\
& \wedge \left(  c_1  = 5 \vee  c_2  = 5 \vee c_3  = 5 \vee c_4  = 5 \vee  c_5
	= 5 \vee  c_6  = 5 \vee  c_7  = 5 \vee  c_8  = 5 \vee  c_9 = 5 \right)\\
& \wedge \left(  c_1  = 6 \vee  c_2  = 6 \vee c_3  = 6 \vee c_4  = 6 \vee  c_5
	= 6 \vee  c_6  = 6 \vee  c_7  = 6 \vee  c_8  = 6 \vee  c_9 = 6 \right)\\
& \wedge \left(  c_1  = 7 \vee  c_2  = 7 \vee c_3  = 7 \vee c_4  = 7 \vee  c_5
	= 7 \vee  c_6  = 7 \vee  c_7  = 7 \vee  c_8  = 7 \vee  c_9 = 7 \right)\\
& \wedge \left(  c_1  = 8 \vee  c_2  = 8 \vee c_3  = 8 \vee c_4  =  8 \vee  c_5
	= 8 \vee  c_6  = 8 \vee  c_7  = 8 \vee  c_8  = 8 \vee  c_9 = 8 \right)\\
& \wedge \left(  c_1  = 9 \vee  c_2  = 9 \vee c_3  = 9 \vee c_4  = 9 \vee  c_5
	= 9 \vee  c_6  = 9 \vee  c_7  = 9 \vee  c_8  = 9 \vee  c_9 = 9  \right)\\
\Leftrightarrow & \bigwedge^9_{n=1}  \bigvee^9_{i=1} \,\,\, c_i = n
\end{align*}
\hspace{0cm}\\
\noindent
This definition follows by the pigeon hole principle. Our aim is to check if the
nine inputs of \texttt{is\_valid} contain each one digit in
$\{1,2,3,4,5,6,7,8,9\}$. Each of the OR-clauses enclosed in brackets check if
each digit is present at least one time, but since we have 9 clauses it follows
that if even a single duplicate digit is given then at least an OR-clause will
have false value. If we imagine each OR-clause is a ``hole'' in which we deposit
digit, in this case the ``hole'' corresponding to the duplicate digit will be
``fed'' twice, and since we have only 9 ``holes'' and 9 ``pigeons'', the
duplicate will cause a ``hole'' to be void of ``pigeons'', thus making the
entire \texttt{is\_valid} formula false and correctly rejecting the invalid
input.
\hspace{0cm}\\

@@

\setcounter{equation}{0}
$\text{sudoku}\left( \{ c_{i,j} \}_{i, j \in \{ 1,2,3,4,5,6,7,8,9\}} \right) = $\begin{align}
&\bigwedge^9_{i=1} \text{ is\_valid}\left(  c_{i,1},  c_{i,2},  c_{i,3},  c_{i,4},  c_{i,5},  c_{i,6},  c_{i,7},  c_{i,8},  c_{i,9}  \right)\\
\wedge &\bigwedge^9_{j=1} \text{ is\_valid}\left( c_{1,j},  c_{2,j},  c_{3,j},  c_{4,j},  c_{5,j},  c_{6,j},  c_{7,j},  c_{8,j},  c_{9,j} \right) \\
	\wedge &\bigwedge^9_{i,j \in \{1,\,4,\,7\}} \text{ is\_valid}\left( c_{i,j},
	c_{i,(j+1)},  c_{i,(j+2)}, c_{(i+1),j},  c_{(i+1),(j+1)},
	c_{(i+1), (j+2)}, c_{(i+2),j},  c_{(i+2),(j+1)},  c_{(i+2),(j+2)} \right).
\end{align}
\newpage
\section{Development}\label{development}
We have developed the project in Python3 to provide both a web interface and the solver logic using z3.\\
For what concerns the z3 SAT solver, we have used the official \texttt{z3–solver} Python3 library to write the z3 logic instead of writing a z3 \texttt{.dimacs} file for every execution and invoking the z3 solver using a system shell. This allows for better performance and easier maintenance of the code.
The \texttt{z3–solver} Python3 library allows to specify constraints programmatically and is able to provide both information about the satisfiability information and data structure models that represent a valid solution for the problem.\\ \\
First of all, we want to translate the following definition into Python3 \texttt{z3–solver} code:
\begin{align*}
\text{sudoku}\left( \{ c_{i,j} \}_{i, j \in \{ 1,2,3,4,5,6,7,8,9 \}} \right) = & \nonumber\\
       & \bigwedge^9_{i=1} \text{ is\_valid}\left(  c_{i,1},  c_{i,2},  c_{i,3},  c_{i,4},  c_{i,5},  c_{i,6},  c_{i,7},  c_{i,8},  c_{i,9}  \right)\\
\wedge &\bigwedge^9_{j=1} \text{ is\_valid}\left( c_{1,j},  c_{2,j},  c_{3,j},  c_{4,j},  c_{5,j},  c_{6,j},  c_{7,j},  c_{8,j},  c_{9,j} \right) \\
\wedge &\bigwedge^9_{i,j \in B} \text{ is\_valid}\left( c_{i,j},  c_{i,(j+1)},  c_{i,(j+2)},  c_{(i+1),j},  c_{(i+1),(j+1)},  c_{(i+1),(j+2)},  c_{(i+2),j},  c_{(i+2),(j+1)},  c_{(i+2),(j+2)} \right).
\end{align*}

@@

\begin{align*}
\text{is\_valid}\left(  c_1,  c_2,  c_3,  c_4,  c_5,  c_6,  c_7,  c_8,  c_9
	\right) \Leftrightarrow \bigwedge^9_{n=1}  \bigvee^9_{i=1} \,\,\, c_i = n
\end{align*}
